3D MAGNETIC MODELLING FOR ELIPSOIDS

### 2016-10-06 manuscript/Paper-magnetic-elipsoid_metodologia.tex
paper-elipsoid modificacao metodologia do elipsoide

\title{3D MAGNETIC MODELLING FOR ELLIPSOIDS}

@@

\begin{abstract}
In this work we will present results from a numerical modeling of the magnetic field and total field anomaly, represented by prolate and triaxial ellipsoids sources. Such approach provide analytical results for anisotropy of magnetic susceptibility as well as for self-demagnetization effects, which can be easily adapted for distinctive geologic structures - hence being an useful tool for educational (e.g., potential methods and rock magnetism) and applied geophysics (e.g., characterization of high magnetic susceptibility, mineralized bodies) purposes. Numerical tests by means of a Python code (currently under development) allowed us to compared the effects of different geometries (ellipsoidal sources, spheres, dipole lines and elliptic cylinders), which were used to validate our computational implementation. This code will be freely available to the scientific community by the end of the year.
\end{abstract}

@@

The magnetic method is considered one of the oldest geophysics techniques. The greek philosofer Thales supposedly did the first obsevations on magnetism in the sixth century. The chineses in the XII century, already used magnetic compasses as an instrument for marine orientation \citep{nabighian2005historical}. During the World War II there was a need for instruments of several magnitudes of precision higher than was available at the time for detection of submarines by aeromagntic exploration. Thus, the fluxgate magnetometer was invented, expanding the use of this method for geophysic exploration \citep{reford1964aeromagnetics,hanna1924some}. The diffusion of this method was quick and well accepted whereas the possiblity  of covering large areas in a relativey short time \citep{blakely1996potential,nabighian2005historical}.

The applications of magnetic exploration includes several important study situations as estimating the average of basement relief, mapping geologic structures like faults and lithologic contacts; and also for mineral exploration, identifying mineral deposits and mapping geologic traps for oil and gas  \citep{oliveira2015estimation}.

In the decade of 1970 \cite{farrar1979some} showed in his work the value of the ellipsoidal model in geophysic exploration, where it was used for analysis of magnetic anomalies of Tennant Crees's gold mine in Australia. The ellipsoid demonstrated to be the most competent geometry to model bodies of lenticular pipe shape and the most appropriate to handle self-desmagnetization of bodies of high susceptibilty - magnetic susceptibility higher than 0.1 SI \citep{clark2014methods}.

In his book \textit{A Treatise on Electricity and Magnetism}, \cite{maxwell1881treatise} demonstrated that only shapes bounded by second degree surfaces are uniformly magnetized when placed in a uniform field. In the particular case of the ellipsoidal geometry, the intern magnetic field is independent of spatial coordinates, therefore its magnetization is completely homogeneous, making it the only geometric shape that has a true anlitic solution for self-desmagnetization \citep{clark1986magnetic}.

The magnetic field resultant of an ellipsoid was presented the first time in full modeling in the space domain by \cite{emerson1985magnetic}, along a compilation of several others geometric shapes. Through a consistent notation this work tapped a hole in the literature that missed the ellipsoidal model. This model, however, consists in a division between prolate and oblate ellipsoids, with some differences in the algorithm.

In this work, we will implement a generalized model for triaxial ellipsoids published by \cite{clark1986magnetic}. From the solution of gravitational potential for uniform ellipsoids, formally solved by \cite{dirichlet1839nouvelle} in his article \textit{Sur un nouvelle methode pour la determination des integrales multiples}, that uses first and second Legendre's normal elliptic integrals, and using the Poisson's relation \citep{grant1965interpretation} it is possible to calculate the magnetic potential, and posteriorly the magnetic field generated by the body calculating the gradient of this potential.

\section{Methodology}
The total-field anomaly can be described as a difference between the total field vector and the induced magnetization of the crust of the Earth, thats includes any anomalous distribuition of magnetization that can occur by magnetizied bodies in subsurface. This difference can be written as:
\begin{equation} \label{eq:1}
\Delta T^0_i = ||T_i|| - ||F_i||,
\end{equation}
where $\Delta T^0_i$ is the observed vector of total-field anomaly in the $i$-th position ($x_i$, $y_i$, $z_i$), $i = 1,...N$, $F_i$ is the geomagnetic field and $||.||$ is the Euclidian norm. The total-field vector $T_i$ is:
\begin{equation}\label{eq:2}
T_i = F_i + B_i,
\end{equation}
and $B_i$ is the total induced magnetization vector produced by all sources of anomalous susceptibility distribuition.
Considering $F_i$ a constant vector, $F_0$ for local ou regional scale, and that $||F_0|| \gg ||B_0||$, since, $B_i$ is a small pertubation of the magnetic field $F_i$, we can approximate the euclidian norm of the vector $T_i$ by a Taylor's expansion of first order:
\begin{equation}\label{eq:3}
||T_i|| \approx ||F_0 + B_i|| \approx ||F_0|| + F^T B_i
\end{equation}
where $T$ indicates transposition and
\begin{equation}\label{eq:4}
\hat{F} = \dfrac{F_o}{||F_0||}
\end{equation}

@@

\subsection{The Forward model of a triaxial elipsoid}
The implementation of the forward problem of a triaxial ellipsoid's magnetic field (three semi-axis $a > b > c$) is done in a new coordinate system, where its center is the center of this ellipsoidal body.

We will define this new coordinate system ($x_1,x_2,x_3$) by the unit vectors $\hat{v}_h (h=1,2,3)$ with the respect to the geographic axis $x$, $y$ e $z$:
\begin{equation}
\hat{v_1} = (l_1,m_1,n_1) = (-\cos\alpha \cos\delta, -\sin\alpha \cos\delta, -\sin\delta)
\end{equation}

@@

The angles refering to the unit vectors are determined by the orientations of the elipsoid's semi-axis. The angle $\alpha$ is the azimuth of semi-major axe ($a$) plus 180$^{\circ}$. Meanwhile $\delta$ is the inclination of semi-major axe ($a$) in relation to the geographic plane. Lastly, $\gamma$ is the angle between the semi-mid axe ($b$) and the vertical projection of the ellipsoid's center with the geographic plane.

@@

Where $x_c$, $y_c$ e $z_c$ are the coordinates of the ellipsoid's center in the geographic system $x$, $y$ e $z$.

For an ellipsoid of semi-axis $a > b > c$, the equation that defines yours surface is:

@@

The parameter $s$ controls the ellipsoid form. When $s$ gets close to $\infty$ the equation \eqref{eq:elipsoide} tends to the sphere equation of radius $r=\sqrt{\lambda}$. When $s = -c^2$, the last term of the ellipsoid's equation is less than zero and she becomes the equation of a circle.

@@

This set of roots, called ellipsoidal coordinates, correspond to parameters of a point ($x_1,x_2,x_3$) which are under the intersection of three ortogonal surfaces related to the body coordinates. Their expressions are:

@@

Where $F(\theta,k)$ e $E(\theta,k)$ are first and second Legendre's normal elliptical integrals, respectively. To calculate $k$ and $\theta$ we have the following expressions:

@@

Note that $L_r, M_r, N_r$ ($r = 1,2,3$), $l,m,n$ e $l_N, m_N, n_N$ depends on their respective vector's inclinations and declinations:
\begin{equation}
L_R,l,l_N = \cos(dec) \, \cos(inc)
\end{equation}

@@

This way, the components of the magnetic field produced by a triaxial ellipsoid in the body's coordinates outside of it is:

@@

Both $F(\theta^{'}, k)$ as $E(\theta^{'}, k)$ are again the first and second  noraml Legendre's elliptical integrals. While $A(\lambda), B(\lambda), C(\lambda)$ are analytic solutions of the integrals of the potential equation of an ellpsoid. This problem was solved by Dirichlet in 1839 \citep{clark1986magnetic} for the gravitational potential given by:

@@

Using the equations from \eqref{eq:campo_e} to \eqref{eq:integralC}, we can rearrange them in a matrix format that will be more computationally efficient and will help us better see the inverse problem ahead. Thus, the magnetic field generated by the ellipsoid is:

@@

For $j = 1,...,L$, the number of ellipsoids been modeled and $i = 1,...,N$  for the $i$th element of the calculated field. The total magnetic field then is:

>>>>

\title{3D MAGNETIC MODELLING FOR ELIPSOIDS}

@@

\begin{abstract}
In this work we will present results from a numerical modeling of the magnetic field and total field anomaly, represented by triaxial and prolate elipsoidal sources. Such approach provide analytical results for anisotropy of magnetic susceptibility as well as for self-demagnetization effects, which can be easily adapted for distinctive geologic structures - hence being an useful tool for educational (e.g., potential methods and rock magnetism) and applied geophysics (e.g., characterization of high magnetic susceptibility, mineralized bodies) purposes. Numerical tests by means of a Python code (currently under development) allowed us to compared the effects of different geometries (ellipsoidal sources, spheres, dipole lines and elliptic cylinders), which were used to validate our computational implementation. This code will be freely available to the scientific community by the end of the year.
\end{abstract}

@@

The magnetic method is considered one of the oldest geophysics techniques. The greek philosofer Thales supposedly did the first obsevations on magnetism in the sixth century. The chineses in the XII century, already used magnetic compasses as an instrument for marine orientation \citep{nabighian2005historical}. During the World War II there was a need for instruments of several magnitudes of precision higher than was available at the time for the detection of submarines trough aeromagnetic exploration. Thus, the fluxgate magnetometer was invented, expanding the use of this method for geophysic exploration \citep{reford1964aeromagnetics,hanna1924some}. The diffusion of this method was quick and well accepted whereas the possiblity  of covering large areas in a relatively short time \citep{blakely1996potential,nabighian2005historical}.

The applications of magnetic exploration includes several important study situations, as estimating the average of basement relief, mapping geologic structures as faults and lithologic contacts; and also for mineral exploration, identifying mineral deposits and mapping geologic traps for oil and gas  \citep{oliveira2015estimation}.

In the decade of 1970 \cite{farrar1979some} showed in his work the value of the elipsoidal model in geophysic exploration, where it was used for analysis of magnetic anomalies in Tennant Creeks's gold mine in Australia. The elipsoid was the most correct geometry to model bodies of lenticular pipe shape and the most appropriate to handle self-desmagnetization of bodies of high susceptibilty - magnetic susceptibility higher than 0.1 SI \citep{clark2014methods}.

In his book \textit{A Treatise on Electricity and Magnetism}, \cite{maxwell1881treatise} demonstrated that only shapes bounded by second degree surfaces are uniformly magnetized when placed in a uniform field. In the particular case of the elipsoidal geometry, the intern magnetic field is independent of spatial coordinates, therefore its magnetization is completely homogeneous, making it the only geometric shape that has a true analitic solution for self-desmagnetization \citep{clark1986magnetic}.

The magnetic field resultant of an elipsoid was presented the first time in full modeling in the space domain by \cite{emerson1985magnetic}, along a compilation of several others geometric shapes. Through a consistent notation this work tapped a hole in the literature that missed the elipsoidal model. This model, however, consists in a division between prolate and oblate ellipsoids, with a few differences in the algorithm.

In this work, we will implement a generalized model for triaxial ellipsoids published by \cite{clark1986magnetic}. From the solution of gravitational potential for uniform elipsoids, formally solved by \cite{dirichlet1839nouvelle} in his article \textit{Sur un nouvelle methode pour la determination des integrales multiples}, that uses first and second Legendre's normal elliptic integrals, and using the Poisson's relation \citep{grant1965interpretation} it is possible to calculate the magnetic potential, and posteriorly the magnetic field generated by the body calculating the gradient of this potential.

\section{Methodology}
The total-field anomaly can be described as a difference between the total field vector and the induced magnetization of the crust of the Earth, that includes any anomalous distribuition of magnetization that can occur by magnetizied bodies in subsurface. This difference can be written as:
\begin{equation} \label{eq:1}
\Delta T^0_i = ||T_i|| - ||F_i||,
\end{equation}
where $\Delta T^0_i$ is the observed vector of total-field anomaly in the $i$-th position ($x_i$, $y_i$, $z_i$), $i = 1,...N$, $F_i$ is the geomagnetic field and $||.||$ is the Euclidian norm. The total-field vector $T_i$ is:
\begin{equation}\label{eq:2}
T_i = F_i + B_i,
\end{equation}
and $B_i$ is the total induced magnetization vector produced by all sources of anomalous susceptibility distribuition.
Considering $F_i$ a constant vector, $F_0$ for local ou regional scale, and that $||F_0|| \gg ||B_0||$, since, $B_i$ is a small pertubation of the magnetic field $F_i$, we can approximate the euclidian norm of the vector $T_i$ by a Taylor's expansion of first order:
\begin{equation}\label{eq:3}
||T_i|| \approx ||F_0 + B_i|| \approx ||F_0|| + F^T B_i
\end{equation}
where $T$ indicates transposition and
\begin{equation}\label{eq:4}
\hat{F} = \dfrac{F_o}{||F_0||}
\end{equation}

@@

\subsection{The Forward model of a triaxial elipsoid}
The implementation of the forward problem of a triaxial elipsoid's magnetic field (three semi-axis $a > b > c$) is done in a new coordinate system, where its origin is the center of this elipsoidal body.

This new coordinate system ($x_1,x_2,x_3$) is defined by the unit vectors $\hat{v}_h (h=1,2,3)$ with the respect with the geographic axis $x$, $y$ e $z$:
\begin{equation}
\hat{v_1} = (l_1,m_1,n_1) = (-\cos\alpha \cos\delta, -\sin\alpha \cos\delta, -\sin\delta)
\end{equation}

@@

The angles referring to the unit vectors are determined by the orientations of the elipsoid's semi-axis. The angle $\alpha$ is the azimuth of semi-major axe ($a$) plus 180$^{\circ}$. While $\delta$ is the inclination of semi-major axe ($a$) in relation to the geographic plane. Lastly, $\gamma$ is the angle between the semi-mid axe ($b$) and the vertical projection of the elipsoid's center with the geographic plane.

@@

Where $x_c$, $y_c$ e $z_c$ are the coordinates of the elipsoid's center in the geographic system $x$, $y$ e $z$.

For an elipsoid of semi-axis $a > b > c$, the equation that defines your surface is:

@@

The parameter $s$ controls the elipsoid form. When $s$ gets close to $\infty$ the equation \eqref{eq:elipsoide} tends to the sphere equation of radius $r=\sqrt{\lambda}$. When $s = -c^2$, the last term of the elipsoid's equation is less than zero and it becomes the equation of a circle.

@@

This set of roots, called elipsoidal coordinates, correspond to the parameters of a point ($x_1,x_2,x_3$) which are under the intersection of three ortogonal surfaces related to the body coordinates. Their expressions are:

@@

Where $F(\theta,k)$ e $E(\theta,k)$ are first and second Legendre's normal eliptical integrals, respectively. To calculate $k$ and $\theta$ we have the following expressions:

@@

Note that $L_r, M_r, N_r$ ($r = 1,2,3$), $l,m,n$ e $l_N, m_N, n_N$ depends on their respective vectors of inclinations and declinations:
\begin{equation}
L_R,l,l_N = \cos(dec) \, \cos(inc)
\end{equation}

@@

This way, the components of the magnetic field produced by a triaxial elipsoid in the body's coordinates outside of it is:

@@

Both $F(\theta^{'}, k)$ as $E(\theta^{'}, k)$ are again the first and second  noraml Legendre's eliptical integrals. While $A(\lambda), B(\lambda), C(\lambda)$ are analytic solutions of the integrals of the potential equation of an elipsoid. This problem was solved by Dirichlet in 1839 \citep{clark1986magnetic} for the gravitational potential given by:

@@

Using the equations from \eqref{eq:campo_e} to \eqref{eq:integralC}, we can rearrange them in a matrix format that will be more computationally efficient. Thus, the magnetic field generated by the elipsoid is:

@@

For $j = 1,...,L$, the number of elipsoids been modeled and $i = 1,...,N$  for the $i$th element of the calculated field. The total magnetic field then is:
